### 2020-10-13 doc/manuscript/Dow_manuscript.Rmd
adding sources

A mixed effect model was used to test the response of wood phenology variables (25% DOY, 50% DOY, 75% DOY, max-rate, max-rate DOY, total growth) to fixed effects of wood-type and Spring temperature, along with random effects of species and tag. We ran two separate models of each wood phenology variable for each of our major wood-types, and one combined model to use for comparison.

@@

•	Bayesian variable results (PRE-JOINT MODEL, updating ASAP)

Ring-porous 25% DOY estimate: -2.23 (credible interval: -2.94 - -1.55)

Ring-porous 50% DOY estimate: - 2.10 (credible interval: -2.71 - -1.42)

Ring-porous 75% DOY estimate: -1.89 (credible interval: -2.53 - -1.28)

Ring-porous total growth estimate: 0.0552 (credible interval: 0.00346 - 0.108)

Ring-porous max-rate DOY estimate: -2.20 (credible interval: -2.77 - -1.62)

Ring-porous max-rate estimate: 0.00113 (credible interval: 0.000151 – 0.00208)

Diffuse-porous 25% DOY estimate: -1.83 (credible interval: -2.86 - -0.822)

Diffuse-porous 50% DOY estimate: -2.39 (credible interval: -3.35 - -1.43)

Diffuse-porous 75% DOY estimate: -2.96 (credible interval: -4.03 - -1.92)

Diffuse-porous total growth estimate: -0.0642 (credible interval: -0.172 – 0.0460)

Diffuse-porous max-rate DOY estimate: -2.54 (credible interval: -3.41 - -1.69)

Diffuse-porous max-rate estimate: 0.00209 (credible interval: 0.000118 – 0.00413)

### Discussion
-Growth timing
We have shown that intraannual growth patterns in North American temperate trees change in response to increasing Spring temperatures. The magnitude of these changes depends largely on the evolutionary differences inherent in the two wood types analyzed here. Ring-porous trees must repair winter embolism of their large xylem vessels before leaf expansion, achieving roughly 25% total yearly growth simultaneously with bud burst, and 50% yearly growth around the time of 75% leaf emergence (D'Orangeville et al,). On the other hand, diffuse-porous trees are less prone to cavitation and so do not need to repair winter embolism before bud burst, allowing up to 75% leaf out before reaching their 25% yearly growth totals (D'Orangeville et al). These differences are likely evolutionary hydraulic adaptations to climate(). The presence of wide-xylem cells in ring-porous trees allows for greater water conductivity in the Spring at the cost of increased cavitation risk during Spring drought (). The pre-leaf stem growth occurring in these species is fueled by carbon stores often used for other cell maintenance activities within the plant (). The dependence on these carbon stores for cambial activation could prove to be a detriment to the tree, as exhaustion of reserves leads to carbon starvation (). Diffuse-porous trees have narrower xylem cells lowering the risk of cavitation, but leading to a decreased water conductivity in the Spring, focusing on a faster, shorter growing season in the Summer. This allows the diffuse-porous trees to conserve their carbon stores, making them potentially more resistant to pest/pathogen invasions?, but exposing them to higher temperatures and drought risk during their Summer focused growth period.

@@

Ring-porous trees reach 25% and 50% growth milestones earlier (insert actual # days?) in the year, while 75% remains statistically unchanged, in response to warmer temperautres during early Spring. This implies that the growing season of ring-porous trees expanded due to Spring temperatures, which are projected to continue to increase as anthropogenic global warming intensifies. The reactivation of cambial cells is likely beginning earlier potentially exposing these trees to higher risk of cavitation or leaf damage during frost events (). Damage of this nature would then need to be repaired using carbon reserves which may be in short demand as they are already being used for initial growth. Hypothetically, this could lead to carbon starvation in some trees. Meanwhile, diffuse-porous growth timing seems relatively unaffected by early Spring temperatures, instead showing a higher correlation with late Spring or Summer temps. This disconnect means that diffuse-porous trees will continue tofocus their growing season in the Summer, which is projected to become less condusive to tree growth as temperatures and drought intensity increase. However, diffuse-porous trees are well adapted to drought conditions, and may continue to be successful even in future climate scenarios. If not, and they begin to suffer without being able to adapt by shifting their growing period earlier, the range of some of these species may migrate North to cooler summer conditions becoming potentially extant in the Southern / mid Atlantic of the USA.    (Is mortality of ring-porous species higher in recent decades? Hard to pinpoint, but would likely be caused by warmer early-spring followed by late frost events exhausting NSC reserves?)

This analysis is one of a first generation of intraannual dendroband tree growth studies, and as such, is the metaphorical tip of the ice berg. Here, we focused our analysis purely on early Spring warming. This falls short of the reality of our changing climate, where each season is being affected in some way, leading to a combination of stressors on plant growth. For example, diffuse porous trees seemed to show little, if any, relationship with the climate window defined here, but if we shift that window forward 20 days, the effect increases dramatically, signalling a greater importance of May/June TMAX on trees of that wood type (figure or table?). This revelation agrees with the results from Helcoski et al (2019), which found differences in interspecific growth responses to several climate variables throughout the year using tree cores. Additionally, Other studies have found that Autumn leaf senescence is being delayed potentially in response to warmer temperatures in the previous year's Winter or Autumn (), changes in chilling requirements due to current year Autumn and winter temperatures (), and stunted total productivity induced by drought and high TMAX in the current year's Summer (). How these relationships may impact intraannual growth patterns or variables like those described here is uncertain, but identifying such relationships will help us further understand how climate change is impacting forests.

We present a mere stepping stone for future studies to build upon.

Diffuse-porous 25-75% growth period shrinking leaving trees more vulnerable to drought while ring-porous 25-75% growing, potentially making effect of drought less damaging? But RP still prone to cavitation + longer growth period could mean exposure to higher number of droughts resulting in serious damage to water transport system?

DIfferences in total relationship could be because of response difference in wood types OR could be because the diffuse-porous’ growth occuring during the later summer means it is exposed to drought/heat and on average, not feeling the effects of the earlier spring. differences in wood porosity growth timing means exposure to different periods of climate stress ring porous adaptable to spring temps, diffuse not so much leads to higher growth in ring however, diffuse still more resistant to cavitation potential for more frequent late frost events having large impacts as spring warms? offset of chilling requirements and spring leaf out requirements

much assumed about future forest productivity based on gis different wood types/species may respond differently need to understand at a finer scale what is happening

>>>>

A mixed effect model was used to test the response of wood phenology variables (25% DOY, 50% DOY, 75% DOY, max-rate, max-rate DOY, total growth) to fixed effects of wood-type and Spring temperature, along with random effects of species and tag. We ran two separate models of each wood phenology variable for each of our major wood-types, and one combined model to use for comparison. Models were run using the LME4 package in R (Bates, D., Mächler, M., Bolker, B. & Walker, S. (2015). Fitting linear mixed‐effects models using lme4. J. Stat. Softw., 67, 1–48.)

@@

### Discussion
-Growth timing
We have shown that intraannual growth patterns in North American temperate trees change in response to increasing Spring temperatures. The magnitude of these changes depends largely on the evolutionary differences inherent in the two wood types analyzed here. Ring-porous trees must repair winter embolism of their large xylem vessels before leaf expansion, achieving roughly 25% total yearly growth simultaneously with bud burst, and 50% yearly growth around the time of 75% leaf emergence (D'Orangeville et al,). On the other hand, diffuse-porous trees are less prone to cavitation and so do not need to repair winter embolism before bud burst, allowing up to 75% leaf out before reaching their 25% yearly growth totals (D'Orangeville et al). These differences are likely evolutionary hydraulic adaptations to climate(). The presence of wide-xylem cells in ring-porous trees allows for greater water conductivity in the Spring at the cost of increased cavitation risk during Spring drought, and high likeliehood of damage during frost events (Cochard & Tyree, 1990). The pre-leaf stem growth occurring in these species is fueled by carbon stores, often used for other cell maintenance activities within the plant (). The dependence on these carbon stores for cambial activation could prove to be a detriment to the tree, as exhaustion of reserves leads to carbon starvation (). Diffuse-porous trees have narrower xylem cells lowering the risk of cavitation, but leading to a decreased water conductivity in the Spring, revealing a growth strategy more focused on a relatively faster, shorter growing season in the Summer. This allows the diffuse-porous trees to conserve their carbon stores, making them potentially more resistant to pest/pathogen invasions?, but exposing them to higher temperatures and drought risk during their Summer-centered growth period.

@@

Ring-porous trees reach 25% and 50% growth milestones earlier (insert actual # days?) in the year, while 75% remains statistically unchanged, in response to warmer temperatures during early Spring. This implies that the growing season of ring-porous trees expanded due to Spring temperatures, which are projected to continue to increase as anthropogenic global warming intensifies. The reactivation of cambial cells is likely beginning earlier potentially exposing these trees to higher risk of cavitation or damage during frost events (Cochard & Tyree, 1990). Damage of this nature would then need to be repaired using carbon reserves which may be in short demand as they are already being used for initial growth. Hypothetically, this could lead to carbon starvation in some trees. Alternatively, the extension of growth season and higher yearly growth totals like we see here, could give ring-porous trees a competitive advantage over their non-responsive diffuse-porous counterparts. In the long term, this could lead to changes in species dominance within temperate forests (talk more aobut this?).

Meanwhile, diffuse-porous growth timing seems relatively unaffected by early Spring temperatures, instead showing a higher correlation with late Spring or Summer temps. This disconnect means that diffuse-porous trees will continue to focus their growing season in the Summer, which is projected to become less conducive to tree growth as temperatures and drought intensity increase. However, diffuse-porous trees are adapted to survive in drought conditions, and may continue to be successful even in future climate scenarios (RIght? seeing conflicting reports. Sources.). Studies have found that these diffuse-porous trees in the Eastern USA show the largest reduction in ring-width increment during, and following, Summer droughts (Kannenberg et al, 2018) as well as being generally more sensitive to Summer precipitation patterns (Elliot et al, 2015). The potential for more frequent reduction in growth coupled with the increased growing season length of ring-porous species may lead to an alteration in forest species composition.  --the range of some of these species may migrate North to cooler summer conditions becoming potentially extant in the Southern / mid Atlantic of the USA.--    (Is mortality of ring-porous species higher in recent decades? Hard to pinpoint, but would likely be caused by warmer early-spring followed by late frost events exhausting NSC reserves?)

This analysis is one of a first generation of intraannual dendroband tree growth studies, and as such, is the metaphorical tip of the ice berg. Here, we focused our analysis purely on early Spring warming. This falls short of the reality of our changing climate, where each season is being affected in some way, leading to a combination of stressors on plant growth. For example, diffuse porous trees seemed to show little, if any, relationship with the climate window defined here, but if we shift that window forward 20 days, the effect increases dramatically, signalling a greater importance of May/June TMAX on trees of that wood type (figure or table?). This revelation agrees with the results from Helcoski et al (2019), which found differences in interspecific growth responses to several climate variables throughout the year using tree cores. Additionally, Other studies have found that Autumn leaf senescence is being delayed potentially in response to warmer temperatures in the previous year's Winter or Autumn (), changes in chilling requirements due to current year Autumn and winter temperatures (), and stunted total productivity induced by drought and high TMAX in the current year's Summer (), diffuse porous trees increased growth during the following growing season when winter temperatures (TMIN) were lower (Benito & Pederson, 2015; Elliot et al, 2015). How these relationships may impact intraannual growth patterns or variables like those described here is uncertain, but identifying such relationships will help us further understand how climate change is impacting forests.

We present a stepping stone for future studies to build upon.
